Towards a standardized evaluation of imputation methodology

### 2022-09-15 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
Resubmit changes

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policymaking, demonstrated e.g. by the detailed manual created by the National Research Council \citep*{little2012prevention}. Although top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers and data scientists are embracing imputation techniques. After all, the principle of imputation is very intuitive. The idea behind imputation is to impute (fill in) missing values, to obtain a valid estimate of what could have been. The completed data that are thus obtained can be analyzed by standard techniques, which conveniently separates the missing data problem from the analysis.

For inferential analyses, multiple imputation \citep{rubi76} has proven to be a valuable technique for obtaining valid inferences on incomplete datasets. In the case of multiple imputation each missing datum is imputed more than once, leading to multiple completed data sets. The resulting multiple inferences can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. When prediction is the goal of the analysis effort, a single imputation may be sufficient \citep{nijm20}. The quality of a solution obtained by imputation depends on the statistical properties of the incomplete data and the degree to which an imputation procedure is able to capture these properties when modeling missing values. In general, it holds that modeling missing data becomes more challenging when the amount of missingness increases. However, when (strong) relations in the data are present, the observed parts can hold great predictive power for the models that estimate the missingness. In that case, imputation would be substantially more efficient than the ubiquitous complete case analysis.

@@

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. Such studies are typically composed of an analysis pipeline in which data are generated, made incomplete, and imputed repeatedly under different simulation conditions. A set of evaluation criteria is then postulated to evaluate the performance of one or more missing data methods. However, there is no consensus on how to design, execute and report simulation studies aimed to evaluate imputation methodology. Without a `gold standard' for the evaluation of imputation routines, the validity and comparability of simulation setups may differ tremendously from one developer to another. This brings forth a chain of potential problems in the objective assessment of imputation method performance within and across studies, which may lead to sub-optimal use of imputation in practice.

@@

The purpose of this paper is threefold: First, to raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as a reviewer for statistical journals. Second, to provide imputation methodologists with a suggested course of action when using simulation studies to evaluate imputation techniques for missing data problems. This suggested approach should identify common ground but is in no way intended as an absolute solution. This identifies the third purpose of this paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties.

Standardizing the evaluation of imputation methodology requires simulators to consider common aspects in their simulation workflows. Table \ref{table:check} outlines some suggested steps to adopt. Please note that there is--of course--no `silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed structure for planning and reporting simulation studies and \citet{pawe22}'s advice to pre-register simulation protocols. We henceforth highlight and add some recommendations for the evaluation of imputation methodology. 
% [TODO: Add advice to pre-register simulation protocols, which allows for fair comparisons of per-protocol results \citep{pawe22}.]

@@

The ideal evaluation of imputation methodology is study specific. A simulation study developed for the comparison of several imputation methods will typically have a different design than one aimed at establishing the inferential validity of a single (novel) method. How the evaluated methods are used in practice thereby governs the simulation procedure. Since the simulation aims and setup are intertwined with the choice of imputation methods under evaluation, the suitability of the evaluations should be assessed with the imputation methods' purpose in mind.

@@

One potential reason for discrepancies between imputation method performance in simulations versus applications is the use of sampling variance in the simulation design. Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation \citep{vink14}. If sampling variance is not of interest, a single complete data set can serve as comparative truth in all simulation repetitions (as opposed to a `standard' simulation setup, in which data is drawn every repetition). After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Removing sampling variance from performance evaluation is computationally convenient and may potentially sharpen comparisons between imputation methods, clarifying inconclusive results on their ability to correct for missingness. However, when deviating from the standard simulation workflow, conventional pooling rules for multiple imputation \citep[cf.][p. 76-77]{rubi87} do not apply and alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}. This opens the door towards human error in the implementation and interpretation of such a simulation study. On the one hand, we highly value the additional flexibility that this design offers. Especially in the case of data transformations, it can be challenging to derive the true parametric references to evaluate against. Simply using a single generated complete set in which missingness is induced, and evaluating against that generated data set avoids a plethora of procedural problems and computational challenges. On the other hand, we want to warn simulators against potential mishaps, e.g. when calculating and comparing standard errors.

@@

A well-designed simulation does not guarantee fair evaluations. Still, we propose to consider the following simulation design choices. Estimands and other simulation targets should be clearly and unambiguously defined in the context of the study aim \citet{pete14}. The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error][]{morr18}. The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There may, for example, be interactions between the source and amount of missingness in the incomplete data and the efficacy of imputation methods (i.e., the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions) [TODO: add ref dance of the mechanisms!]. If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons.

@@

To evaluate the ability of an imputation routine to handle missingness, a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand by choosing a data-generating mechanism. Data-generating mechanisms define how a complete dataset is obtained at the start of each simulation repetition. There are two general approaches to generating complete data: (i) model-based simulation, in which data are drawn from a known statistical model or probability distribution, such as the multivariate normal distribution; and (ii) design-based simulation, where data are sampled (with replacement) from a sufficiently large observed set, such as official registers. The simulator should choose their data-generating mechanism(s) in line with the study scope. Model-based data-generating mechanisms have advantages in flexibility and precision, since data are generated from a known statistical model and the true theoretical parameters can be derived. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest.

The problem with model-based data-generating mechanisms is that a method's performance on simulated data may not translate to empirical data. Real-life data hardly ever follow a given theoretical distribution, so there is no guarantee that simulation results are generalizable. Moreover, data are often generated such that the problem being studied is most pronounced, e.g. with consistently high correlations between groups of variables. This results in simulated data that contain such valuable information structures that, no matter what type of missingness would subsequently be induced, the observed parts of the data will still hold much (if not all) of the information about the missing part. Unsurprisingly, the performance of any imputation method will then be evaluated as good. Another threat to the generalizability of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the variable type(s), and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.

>>>>

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policymaking, demonstrated e.g. by the detailed manual created by the National Research Council \citep*{little2012prevention}. Although top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers and data scientists are embracing and developing imputation methodology. After all, the principle of imputation is very intuitive. The idea behind imputation is to impute (fill in) missing values to obtain a valid estimate of what could have been. The completed data that are thus obtained can be analyzed by standard techniques, which conveniently separates the missing data problem from the analysis.

For some data analyses a single imputation may suffice, while for other data analyses multiple imputations are needed. For example, in the case of inferential analyses, such multiple imputations \citep{rubi76} have proven to be a valuable technique for obtaining valid inferences on incomplete datasets. The resulting multiple inferences on multiple completed data sets can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. When generating predicted values is the goal of the data analysis, a single imputation may already generate sufficient \citep{nijm20} predicted values.

In all scenarios it can be said that the quality of the solution obtained by imputation depends on the statistical properties of the incomplete data and the degree to which the imputation procedure is able to capture these properties when modeling missing values. In general, it holds that modeling missing data becomes more challenging when the amount of missingness increases. However, when (strong) relations in the data are present, the observed parts can hold great predictive power for the models that estimate the missingness. In that case, imputation would be substantially more efficient than the ubiquitous complete case analysis.

@@

When evaluating the statistical properties - and thereby the practical applicability - of imputation methodology, researchers most often make use of simulation studies. Such studies are typically composed of an analysis pipeline in which data are subsequently generated, made incomplete, and imputed repeatedly under different simulation conditions. A set of evaluation criteria is then postulated to evaluate the performance of one or more missing data methods. Recent attention on best practices for method evaluation by means of simulation \citep{morr18}. and **REF MISSING** \citet{pawe22}'s advice to pre-register simulation protocols will guide methodologist in conceptualizing their simulation workflow. However, there is no consensus on how to design, execute and report simulation studies aimed to evaluate imputation methodology. Without a `gold standard' for evaluation, the validity and comparability of simulation setups may differ tremendously from one developer to another. This brings forth a chain of potential problems in the objective assessment of imputation method performance within and across studies, which could lead to sub-optimal use of imputation in practice.

@@

The purpose of this paper is threefold: First, we raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as reviewers for statistical journals. Second, we provide imputation methodologists with a suggested course of action when using simulation studies to evaluate imputation techniques for missing data problems. This suggested approach should identify common ground, but is in no way intended as an absolute solution. This identifies the third purpose of our paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties.

IK HEB EEN BEETJE MOEITE MET ONDERSTAANDE PARAGRAAF: HET GAAT OM DE REF VAN TIM EN DE PREREG. DIE KUNNEN WE OOK 2 ALINEAS EERDER INFIETSEN. IK HEB EEN POGING GEDAAN.
Standardizing the evaluation of imputation methodology requires simulators adopt common aspects in their simulation and evaluation workflows. Table \ref{table:check} outlines some suggested steps to adopt. Please note that there is--of course--no `silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed structure for planning and reporting simulation studies and \citet{pawe22}'s advice to pre-register simulation protocols. We henceforth highlight and add some recommendations for the evaluation of imputation methodology. 
% [TODO: Add advice to pre-register simulation protocols, which allows for fair comparisons of per-protocol results \citep{pawe22}.]

@@

The ideal evaluation of imputation methodology may differ between studies. A simulation study developed for the comparison of several imputation methods will typically have a different design than one aimed at establishing the inferential validity of a single (novel) method. How the evaluated methods are used in practice thereby governs the simulation procedure. Since the simulation aims and setup are intertwined with the choice of the evaluated imputation methods, one should always asses the suitablity of the chosen evaluation metrics with the imputation methods' purpose in mind.

@@

One potential reason for discrepancies between imputation method performance in simulations versus applications is the use of sampling variance in the simulation design. Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation \citep{vink14}. If sampling variance is not of interest, a single complete data set can serve as comparative truth in all simulation repetitions (as opposed to a `standard' simulation setup, in which data is drawn every repetition). After all, we are interested in solving for the missingness and have no need for the noise induced by the sampling mechanism for evaluation in such studies.

Removing sampling variance from performance evaluation is computationally convenient and may potentially sharpen comparisons between imputation methods, clarifying inconclusive results on their ability to solve for missingness. However, when deviating from the standard simulation workflow, conventional pooling rules for multiple imputation \citep[cf.][p. 76-77]{rubi87} do not apply and alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}. This opens the door towards human error in the implementation and interpretation of such a simulation study. On the one hand, we highly value the additional flexibility that this design offers. Especially in the case of data transformations, it can be challenging to derive the true parametric references to evaluate against. Simply using a single generated complete set in which missingness is induced, and evaluating against that generated data set avoids a plethora of procedural problems and computational challenges. On the other hand, we want to warn simulators against potential mishaps, e.g. when calculating and comparing standard errors.

@@

A well-designed simulation does not guarantee fair evaluations. Still, we propose to consider the following simulation design choices --GV WHICH CHOICES? Ik snap deze eerste zin niet helemaal--. Estimands and other simulation targets should be clearly and unambiguously defined in the context of the study aim \citet{pete14}. The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error][]{morr18}. The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There may, for example, be interactions between the source and amount of missingness in the incomplete data and the efficacy of imputation methods (i.e., the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions) [TODO: add ref dance of the mechanisms!]. If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons.

@@

To evaluate the ability of an imputation routine to handle missingness, a form of ground  truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand by choosing a data-generating mechanism. Data-generating mechanisms define how a complete dataset is obtained at the start of each simulation repetition. There are two general approaches to generating complete data: (i) model-based simulation, in which data are drawn from a known statistical model or probability distribution, such as the multivariate normal distribution; and (ii) design-based simulation, where data are sampled (with replacement) GV IS DIT NODIG? from a sufficiently large observed set, such as official registers. The simulator should choose their data-generating mechanism in line with the study scope. Model-based data-generating mechanisms have advantages in flexibility and precision, since data are generated from a known statistical model and the true theoretical parameters can be derived. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest.

The challenge with model-based data-generating mechanisms is that a method's performance on simulated data may not translate to empirical data. Real-life data hardly ever follow a known theoretical distribution, so there is no guarantee that simulation results are generalizable. Moreover, data are often generated such that the problem being studied is most pronounced, e.g. with consistently high correlations between groups of variables. This results in simulated data that contain such valuable information structures that, no matter what type of missingness would subsequently be induced, the observed parts of the data will still hold much (if not all) of the information about the missing part. Unsurprisingly, the performance of any imputation method will then be evaluated as good.

Another threat to the generalizability of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) GV TEVEEL HAAKJES comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the variable type(s), and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.
